---
title: "Time Series Analysis of Ventura County Wildfires"
subtitle: "Modeling Wildfire Starts and Severity with SARIMA and ARMAX"
author: "Edgar Gonzalez Martinez"
format: html
editor: visual
---

This project analyzes monthly wildfire activity in Ventura County, California from
February 2000 through December 2020. I modeled wildfire starts and wildfire severity
separately, using SARIMA for fire counts and ARMAX for acres burned. The ARMAX model
showed that wildfire severity in Ventura County is meaningfully connected to drought
conditions, Santa Ana-like wind events, and vegetation greenness — not just past fire
behavior.

I chose Ventura County because it is my home county. It also happens to be well-suited
for this kind of analysis: the fire records aggregate cleanly into monthly totals, the
study period captures several major fire seasons, and publicly available environmental
data from NOAA, NASA, and MODIS cover the region well.

[Read the Full Report](https://github.com/CheddyG/Ventura-County-Wildfire-Analysis/blob/main/PSTAT_174-Final_Project_Clean.pdf)

---

## Project Overview

Wildfire counts and acres burned tell different stories. A month with many ignitions
does not necessarily have the most acreage burned — a single wind-driven fire can
dwarf an entire season of smaller ones. Modeling them separately lets each outcome
be explained on its own terms.

**Wildfire starts** were modeled using a seasonal ARIMA approach, capturing temporal
autocorrelation and yearly seasonal structure in the fire count series.

**Wildfire severity** (total acres burned) was modeled using ARMAX — regression with
ARIMA errors — so that environmental predictors could be incorporated while still
accounting for time dependence in the residuals.

---

## Data

The wildfire records come from the USDA Forest Service Fire Program Analysis
Fire-Occurrence Database, filtered to Ventura County and aggregated to monthly totals.
The final sample contains 251 monthly observations.

Environmental predictors for the ARMAX model were pulled from three sources:

- **Drought:** NOAA SPI drought category data (D0–D4), with lags at 1, 3, and 6 months
- **Weather:** NASA POWER daily data aggregated to monthly means — temperature, wind
  speed, relative humidity, and a derived vapor pressure deficit
- **Wind events:** A simplified Santa Ana wind indicator built from daily wind direction,
  maximum wind speed, and relative humidity thresholds
- **Vegetation:** MODIS NDVI (MOD13Q1) averaged over a 20km region around Ventura County

After merging all sources and creating lagged drought variables, the ARMAX dataset
contains 248 complete monthly observations.

---

## Modeling Approach

### Wildfire Starts — SARIMA

Monthly fire counts were log-transformed as `log(1 + fire_count)` to handle the
zero-heavy distribution and reduce the influence of high-count months. Model selection
followed the Box-Jenkins approach: ACF/PACF inspection, differencing tests with
`ndiffs()` and `nsdiffs()`, and comparison of candidate models by AICc and BIC.

The selected model was **ARIMA(0,1,3)(0,0,1)[12]** — one regular difference, three
nonseasonal MA terms, and one seasonal MA term.

### Wildfire Severity — ARMAX

Acres burned was also log-transformed as `log(1 + total_acres_burned)`. Candidate
predictors included all drought lags, temperature, wind speed, humidity, VPD, the
Santa Ana indicator, and NDVI. BIC stepwise selection (with `k = log(n)`) chose three
final predictors: **lagged severe drought (D2 at 3 months), the Santa Ana wind flag,
and mean NDVI.** VIF values were all near 1, so multicollinearity was not a concern.

The ARMAX model was fit using `auto.arima()` with the selected external regressors.
The final error structure was **ARIMA(1,0,1)**.

---

## Results

| Model | AICc | BIC | Ljung-Box p-value |
|---|---:|---:|---:|
| SARIMA — wildfire starts | 467.92 | 485.28 | 0.1224 |
| Baseline SARIMA — acres burned | 1094.54 | 1101.48 | — |
| ARMAX — acres burned | 1065.62 | 1089.66 | 0.8344 |

Both models passed residual diagnostics. The ARMAX model improved on the baseline
SARIMA for acres burned by about 29 AICc points and 12 BIC points. The Ljung-Box
p-value of 0.8344 indicates no meaningful remaining autocorrelation in the ARMAX
residuals.

The ARMAX coefficient signs all align with what fire behavior science would predict:
severe drought at a 3-month lag increases severity, Santa Ana-like wind events increase
severity, and higher vegetation greenness (NDVI) is associated with lower severity —
likely reflecting reduced fuel load or wetter conditions.

---

## Key Visuals

```{r}
#| label: fig-fire-count
#| echo: false
#| warning: false
#| message: false
#| fig-cap: "Monthly wildfire starts in Ventura County, February 2000–December 2020."
#| fig-width: 10
#| fig-height: 5

ggplot(monthly_model, aes(x = month, y = fire_count)) +
  geom_line(linewidth = 0.6) +
  labs(
    title = "Monthly Wildfire Starts in Ventura County",
    x = "Year",
    y = "Monthly wildfire starts"
  ) +
  theme_minimal()
```

```{r}
#| label: fig-annual-acres
#| echo: false
#| warning: false
#| message: false
#| fig-cap: "Annual acres burned in Ventura County."
#| fig-width: 10
#| fig-height: 5

ggplot(annual_model, aes(x = year, y = annual_acres_burned)) +
  geom_col() +
  scale_x_continuous(breaks = seq(2000, 2020, by = 4)) +
  scale_y_continuous(labels = scales::label_number(big.mark = ",")) +
  labs(
    title = "Annual Wildfire Severity in Ventura County",
    x = "Year",
    y = "Annual acres burned"
  ) +
  theme_minimal()
```

---

## Limitations

This analysis works with county-level monthly totals, which smooths over a lot of
local variation in terrain, fuel conditions, ignition sources, and suppression response.
The Santa Ana wind variable is a simplified proxy rather than an official wind event
classification. The NDVI signal reflects general vegetation greenness, not direct fuel
moisture measurements.

Future extensions could include spatial modeling within the county, a more rigorous
wind event dataset, direct fuel moisture data if available, and out-of-sample forecast
evaluation.

---

## Reproducibility

Data sources: USDA Forest Service FPA-FOD, NOAA SPI drought data, NASA POWER,
MODIS MOD13Q1 NDVI. Full code, model diagnostics, and forecast results are in the
report linked above.